\pkg{dataMaid}: your assistant for data cleaning in \proglang{R}

### 2017-02-01 latex/article_vol2.Rnw
Submitted version

\section{Introduction}
Though data cleaning might be regarded as a somewhat tedious activity,
adequate data cleaning is crucial in any data analysis. With
ever-growing dataset sizes and complexities, statisticians and data
analysts find themselves spending a large portion of their time on
data cleaning and data wrangling. While a computer should never
make unsupervised decisions on what should be done to potential
errors in a dataset, it can still be an extremely useful tool in the
data cleaning process. Some errors can be tracked down and flagged by a
computer without further ado, while other types of errors need an analytic
context in order to shine through. But even in the latter case, well-designed
software can aid the process tremendously by giving the human investigator the
information needed for identifying issues.

A number of \proglang{R} packages made for such pre-analysis tools are
already available, including \pkg{janitor} \citep{janitor}, \pkg{plyr}
\citep{plyr}, \pkg{data.table} \citep{data.table}, \pkg{DataCombine}
\citep{DataCombine} , \pkg{validate} \citep{validate}, and
\pkg{DataExplorer} \citep{DataExplorer}. These packages focus on
different stages of the pre-analysis work, most of which are not
really data cleaning.  \pkg{janitor} provides tools for data loading
with a particular emphasis on the challenges of getting neat
\code{data.frame}s from Microsoft Excel data files. \pkg{plyr},
\pkg{data.table} and \pkg{DataCombine} go a few steps further by
providing a wide array of extremely powerful tools for data wrangling,
including a number of particularly useful functions for merging and
working with very large datasets. When it comes to actual data
cleaning, however, the options are fewer.  \pkg{validate} (and the
similar packages \pkg{editrules} and \pkg{deducorrect} from the same
authors) does offer a few tools for identifying and auto-correcting
errors in a dataset, but, as the name suggests, the focus is on
internal validity rather than general data cleaning. In practice, this
means that quite elegant tools for e.g. linear restraints among the
variables in a dataset can be applied, but e.g. looking for
potentially miscoded missing values is not really feasible. The main
difference between these two challenges is the direction in which the
data is inspected: While linear constraints work row-wise with no
ambiguity, determining whether or not something is a miscoded missing
value often requires knowledge about the full variable (e.g. range or
data type), and thus it should be performed
column-wise. \pkg{validate} does not allow for user-defined extensions
of the latter type, thereby limiting its data cleaning potential.
Moreover, no tools have been made for documenting the data cleaning
process.  As automatic cleaning options are also supplied in
\pkg{validate}, we find quite a dangerous cocktail where all power is
given to the the computer with no supervision, neither in real time
nor subsequently.

@@

\begin{table}[htbp]
\footnotesize
\bgroup
\def\arraystretch{1.8}%
\begin{tabular}{L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}}
& \code{summaryFunction} & \code{visualFunction} & \code{checkFunction} \\
\hline
\vspace{0pt} Input (required) & \code{v} - a variable vector \newline \code{...}  &  \code{v} - a variable vector \newline \code{vnam} - the variable name (as character string) \newline \code{doEval} - a logical (\code{TRUE}/\code{FALSE}) controlling the output type of the function & \code{v} - a variable vector \newline \code{nMax} - an integer (or \code{Inf}), controlling how many problematic values are printed, if relevant \newline \code{...}  \\
Input (optional) &  \code{maxDecimals} - number of decimals printed in outputted numerical values.  & - &  \code{maxDecimals}  - number of decimals printed in outputted numerical values.  \\
Purpose & Describe some aspect of the variable, e.g., a central value, its dispersion or level of missingness. & Produce a distribution plot. & Check a variable for a specific issue and, if relevant, identify the values in the variable that cause the issue. \\
Output (required) & A list with entries: \newline \code{\$feature} - a label for the summary value (as character string); \newline \code{\$result} - the result of the summary (as character string) & A character string with \proglang{R} code for producing a plot. This code should be standalone, i.e., should include the data if necessary. & A list with entries: \newline \code{\$problem} - a logical identifying whether an issue was found; \newline \code{\$message} - a character string (possibly empty) decribing the issue that was found, properly escaped and ready for use in \code{rmarkdown} \\
Output (recommended) & A \code{summaryResult} object, i.e., an attributed list with entries \code{\$feature}, \code{\$result} and \code{\$value}, the latter being the values from \code{\$result} in their original format). & \bgroup
\def\arraystretch{1.8}%
\hspace*{-.2cm}\begin{tabular}[t]{p{0.45\linewidth} |
                 p{0.45\linewidth}} \vspace*{-1cm}\raggedright If
                 \code{doEval} is \code{TRUE}: \newline A plot that is
                 opened by the graphic device in \proglang{R}. &
                                                          \vspace*{-1cm}\raggedright If \code{doEval} is \code{FALSE}: \newline   A text string with \proglang{R} code, as described above. \end{tabular} \egroup  & A \code{checkResult} object, i.e., an attributed list with entries \code{\$problem}, \code{\$message} and \code{\$problemValues}, the latter being either \code{NULL} or the problem causing values, as they were found in \code{v}, whichever is relevant. \\
Tools available for producing the function & \code{summaryResult()} & --- & \code{messageGenerator()} \newline \code{checkResult()}
\end{tabular}
\egroup
\caption{Reference information for creating new functions to be used
  as part of the \mbox{summarize-,} visualize-, and check steps. The three
  columns correspond to each of the three function types. %\hl{I think the formatting here is still sort of chaotic. Maybe include horizontal lines everywhere? Is it allowed in JSS? It's impossible to read what's in which row right now.}
}
\label{table.functionTypes}
\end{table}

@@

\subsubsection{Writing a summaryFunction}
As mentioned above, \pkg{dataMaid} provides a dedicated class for
\code{summaryFunction}s. However, this does not imply that they are
particularly advanced or complicated to create; in fact, they are
nothing but regular functions with a particular
input/output-structure. Specifically, they all follow the template
below:
\begin{Verbatim}
mySummaryFunction <- function(v, ...) {
  res <- [result of whatever summary we are doing]
  summaryResult(list(feature = "[Feature name]", result = res))
}
\end{Verbatim}
The last function called here, \code{summaryResult()}, changes the class
of the output, thereby making a \code{print()} method available for it.
Note that \code{v} is a vector and that \code{res} should be either a
character string or something that will be printed as one. In other
words, e.g., integers are allowed, but matrices are not. Though a lot
of different things can go into the \code{summaryFunction} template,
we recommend only using it for summarizing the features of a variable,
and leaving tests and checks for the \code{checkFunction}s (presented
below).

Adhering to the template above is sufficient for using the freshly
made \code{mySummaryFunction()} in \code{clean()}, but we recommend
furthermore adding the new function to the overview of all summary
functions by converting it to a proper \code{summaryFunction}
object. This is done by calling the \code{summaryFunction()} creator with
the user-defined function as the first argument, and additional arguments
\code{description} (an explanatory text which will be added to the attributes of
the function), and \code{classes} (a vector of variable classes the
user-defined function is intended to be applied to, also stored as an
attribute). In other words, a call on the following form should be made:
\begin{Verbatim}
mySummaryFunction <- summaryFunction(mySummaryFunction,
  description = "[Text describing what the summaryFunction does]",
  classes = c([vector of data types that the function is intended
  for]))
\end{Verbatim}
which adds the new function to the output of an
\code{allSummaryFunctions()} call.
% One comment should be devoted to
% the two attributes of a \code{summaryFunction}.
If the \code{description} argument is left unspecified, the name of the
function (which in this case is \code{"mySummaryFunction"}) will be filled in and
stored under the \code{description} attribute. What happens if the \code{classes} argument is
not specified depends on the type of \code{summaryFunction}. If
\code{mySummaryFunction} is constructed as an \code{S3} generic function with associated
methods, the call to \code{summaryFunction()} will automatically produce
a vector of the names of the classes for which the function can be
called. If \code{mySummaryFunction()} is not an \code{S3} generic and
\code{classes} is left unspecified, the attribute will simply be
left empty. Note that the helper function \code{allClasses()} might be useful
for filling out the \code{classes} argument, as it simply lists all
available classes in \pkg{dataMaid}:

@@

\subsubsection{Writing a visualFunction}
\code{visualFunction}s are the functions that produce the figures in a
\pkg{dataMaid} output document. Writing a \code{visualFunction} is slightly
more complicated than writing a \code{summaryFunction}. This follows from
the fact that \code{visualFunction}s need to be able to output standalone
code for plots in order for \code{clean()} to build standalone
\pkg{rmarkdown} files. We recommend using the following structure:
\begin{Verbatim}
myVisualFunction <- function(v, vnam, doEval) {
  thisCall <- call("[the name of the function used to produce the plot]",
    v, [additional arguments for the plotting function])
  if (doEval) {
    return(eval(thisCall))
  } else return(deparse(thisCall))
}
\end{Verbatim}

In this function, \code{v} is the variable to be visualized,
\code{vnam} is its name (which should generally be passed to
\code{title} or \code{main} arguments in plotting functions) and
\code{doEval} controls whether the output is a plot (if \code{TRUE})
or a character string of standalone code for producing a plot (if
\code{FALSE}). Implementing the \code{doEval = TRUE} setting is not
strictly necessary for a \code{visualFunction}'s use in \code{clean()}, but
it makes it easier to assess what visualization options are available,
and obviously, it is crucial for interactive usage of
\code{myVisualFunction()}. In either case, it should be noted that all
the parameters listed above, \code{v}, \code{vnam} and \code{doEval},
are mandatory, so they must be left as is, even if they are not in
use (c.f. Table~\ref{table.functionTypes}).

@@

\subsubsection{Writing a checkFunction}
The last, but perhaps most important, \pkg{dataMaid} function type is
the \code{checkFunction}. These are the functions that flag issues in
the data in the check step and control the overall flow of the data
cleaning process in the precheck stage. A \code{checkFunction} follows
one of two overall structures, depending on the type of check. Either,
it tries to identify problematic values in the variable (as
e.g., \code{identifyMissing()} does), or it performs a check concerning the
variable as a whole (e.g., the functions used for prechecks and the
function \code{identifyNums()}). We present templates for both types of
\code{checkFunction}s below separately, but it should be emphasized that
formally, they belong to the same class.

@@

If problematic values are to be identified, the template from above
should be expanded to follow a slightly more complicated structure:

@@

One comment should be devoted to the helper function,
\code{messageGenerator()}.  This function's sole purpose is aiding
consistent styling of all \code{checkFunction} messages. The function
simply pastes together the \code{problemMessage} and the
\code{problemValues}, with the latter being quoted and sorted
alphabetically. If the \code{nMax} argument to \code{messageGenerator()} is
not \code{Inf}, only the first \code{nMax} problem values will be pasted
onto the message, accompanied by a comment about how many problem
values were left out (if any).  Note that printing quotes in
\pkg{rmarkdown} requires an extensive amount of character escaping, so
opting for \code{messageGenerator()} really is the easiest solution.

@@

\begin{Verbatim}
myProbValCheckFunction <- function(v, nMax, maxDecimals, ...) {
  ...
  problemStatus <- list(problem = problem,
    problemValues = problemValues)

  if (!is.null(problemValues)) {
    problemStatus$problemValues <- round(problemValues, maxDecimals)
  }
  ...
}
\end{Verbatim}
% $

Now, problematic values will be rounded in the outputted message, while they
will still appear in their original format under the entry \code{\$problemValues} %$
in the outputted \code{checkResult}.

@@

We will now build four new functions and show both how they can be
used interactively and how they can be integrated with the \code{clean()}
function. These four new functions are:
\begin{description}
\item[\code{isID()}] A new \code{checkFunction} intended for use in the precheck-stage. This function checks whether a variable consists exclusively of long ($> 10$ characters/digits) entries that are all of equal length, as this might be personal identification codes that we do not wish to print out in the data summary.
\item[\code{mosaicVisual()}] A new \code{visualFunction} that produces mosaic plots. This function will be used in the \textit{visualize} step of \code{clean()}.
\item[\code{countZeros()}] A new \code{summaryFunction} that counts the number of occurrences of the value \code{0} in a variable. This function will be used in the \textit{summarize} step of \code{clean()}.
\item[\code{identifyColons()}] A new \code{checkFunction} that flags variables in which values have colons that appear before and after alphanumerical characters. This is practical for identifying autogenerated interaction effects. This function will be used in the \textit{check} step of \code{clean()}.
\end{description}
These functions are defined in turn below, and afterwards, an example of how they can be called from \code{clean()} is provided.

\subsubsection{isID --- a new checkFunction without problem values}
First, let's define the \code{isID()} function. As this function is not
supposed to list problematic values, it falls within
the category of \code{checkFunction}s represented by
\code{myFullVarCheckFunction()} in the above. We do not particularly wish
to use this function interactively, so we will stick to the minimal
requirements of a \code{checkFunction} used in \code{check()} (see
Table~\ref{table.functionTypes}). The function can then be defined by

<<>>=
isID <- function(v, nMax = NULL, ...) {
  out <- list(problem = FALSE, message = "")
  # Should work for all but logical and Date variables
  if (class(v) %in% setdiff(allClasses(), c("logical", "Date"))) {
    v <- as.character(v)
    lengths <- c(nchar(v))
    # Check that all lengths are the same and greater than 10 characters
    if (all(lengths > 10) & length(unique(lengths)) == 1) {
      out$problem <- TRUE
      out$message <- "Warning: This variable seems to contain ID codes."
    }
  }
  out
}
@

This is essentially all we need to do in order to include this
function as a precheck-function in \code{clean()}, so we will leave it as
is and move on to the next function, namely \code{mosaicVisual()}.

\subsubsection{mosiacVisual --- a new visualFunction}
\proglang{R} contains a function, \code{mosiacplot()}, which produces
mosaic plots. We intend to use this existing high-level plotting
function as part of the new visualization function.  We will define
the new function such that it gets the full \pkg{dataMaid}
functionality. This can be done using the following code which sets up
the call using the existing function \code{mosaicplot()}.

<<>>=
mosaicVisual <- function(v, vnam, doEval) {
  thisCall <- call("mosaicplot", table(v), main = vnam, xlab = "")
  if (doEval) {                     # Return the graphics
    return(eval(thisCall))
  } else return(deparse(thisCall))  # Else return the code for the call
}
@

@@

Even though \code{mosaicVisual()}, as written above, follows the style of a
\code{visualFunction}, it is not yet truly one and therefore, it will not
appear in an \code{allVisualFunctions()} call. In order to get this
functionality, we need to change its object class. This can be done by
writing

@@

\subsubsection{countZeros --- a new summaryFunction}
This \code{summaryFunction} is defined in the following lines of code:

@@

As this function computes an integer (the number of zeros),
there is no difference between the entries \code{\$result} and
\code{\$value}. If, on the other hand, the result had been a character
string, extra formatting might be required in the \code{\$result} entry
(such as escaping of quotation marks), and in this scenario, the two
entries would have differed. As the result is returned as a
\code{summaryResult} object, a printing method is automatically called
when \code{countZeros()} is used interactively:

@@

\subsubsection{identifyColons --- a new checkFunction with problem values}
The last function mentioned above is \code{identifyColons()}. We
define it using the helper function \code{messageGenerator()} to
obtain a properly escaped message, and we use \code{checkResult()} to
make its output print neatly. In the code we use regular expressions
through the \code{gregexpr()} function to identify colons between two
words.

@@

Note, that for \code{checkFunction}s, the function description will
appear in the document produced by \code{clean()} (in the \textit{Data
  cleaning summary} section of the output), so for this type of
function, the change of class is not only done for the sake of the
\code{allCheckFunctions()} output.

\subsubsection{Calling the new summarize/visualize/check functions from clean()}
Now, we are ready to use these new functions in a \code{clean()}
call. The extended \pkg{dataMaid} output document should have the
following modifications, relative to the standard \pkg{dataMaid} output:
\begin{itemize}
\item We want to add the new pre-check function, \code{isID()}, to the already existing pre-checks.
\item We wish to change the plot type for all variables to the new mosaic plot.
\item We want the new summary function, \code{countZeros()}, to be added to the summaries performed on all variable types but \code{Date} and \code{logical}.
\item We want the new check function, \code{identifyColons()}, to be added to the checks performed on \code{character}, \code{factor} and \code{labelled} variables.
\end{itemize}
These options are specified as follows:

@@

Finally, we present a few examples of how to make \pkg{dataMaid}
solve specific issues related to data cleaning. First, we discuss the
challenges related to cleaning large datasets, particularly in terms
of memory use and computation speed. Next, we show how \pkg{dataMaid}
can be used for problem-flagging. Lastly, we discuss how the
\pkg{dataMaid} output document can be included in other \pkg{rmarkdown}
documents as a way to produce clear and concise documentation of a
dataset. %\hl{I feel like there should be more topics here, but I'm all
%  out of ideas...}

\subsection{Cleaning large datasets}
If the dataset becomes very large, the standard use of \code{clean()}
outlined above might not be ideal. If there is a vast number of
variables, creation of the \pkg{rmarkdown} document might be quite
slow, while a large number of observations will generally affect the
rendering time of the document. This issue is especially relevant to
Windows users, where memory usage is not very efficient. The problem is that older,
32-bit versions of Windows generally cannot access all physical RAM,
while newer 64-bit versions of Windows cannot access virtual memory. Therefore,
memory allocation issues becomes even more pressing when using this operating system.

In this section, we give a few
practical examples of ways to deal with large data, while wishing to
still produce (potentially very long) data cleaning overview
documents. Note that the interactive tools of \pkg{dataMaid} can be used
as usual or sequentially in small subsets of the large dataset, if no
such overview documents are needed.

@@

Of course, even less heavy plots might be achieved by writing new
\code{visualFunction}s, using the guidelines from section
\ref{sec:functionTemplates}. For instance, a future extension of
\pkg{dataMaid} might be the inclusion of ASCII plots, as
e.g., represented in the \proglang{R} package \pkg{txtplot} \citep{txtplot}.

@@

Now only the checking results are printed, and only for variables
where problems were identified. An even more minimal output can be
generated by also leaving out the checking results --- then
\code{clean()} essentially just produces a list of the variable names
that should be investigated further:

>>>>

\section{Introduction}
Though data cleaning might be regarded as a somewhat tedious activity,
adequate data cleaning is crucial in any data analysis. With
ever-growing dataset sizes and complexities, statisticians and data
analysts find themselves spending a large portion of their time on
data cleaning and data wrangling. While a computer should never
make unsupervised decisions on what should be done to potential
errors in a dataset, it can still be an extremely useful tool in the
data cleaning process. Some errors can be tracked down and flagged by a
computer without further ado, while other types of errors need an analytic
context in order to be identified. Even in this latter case, well-designed
software can aid the process tremendously by providing the human
investigator with the
information needed for identifying issues.

A number of \proglang{R} packages made for such pre-analysis tools are
already available, including \pkg{janitor} \citep{janitor}, \pkg{plyr}
\citep{plyr}, \pkg{data.table} \citep{data.table}, \pkg{DataCombine}
\citep{DataCombine} , \pkg{validate} \citep{validate}, and
\pkg{DataExplorer} \citep{DataExplorer}. These packages focus on
different stages of the pre-analysis work, most of which are not
really data cleaning.  \pkg{janitor} provides tools for data import
with a particular emphasis on the challenges of getting neat data
frames from Microsoft Excel data files. \pkg{plyr}, \pkg{data.table}
and \pkg{DataCombine} go a few steps further by providing a wide array
of extremely powerful tools for data wrangling, including a number of
particularly useful functions for merging and working with very large
datasets. When it comes to actual data cleaning, however, the options
are fewer.  \pkg{validate} (and the similar packages \pkg{editrules}
\citep{editrules} and \pkg{deducorrect} \citep{deducorrect} from the
same authors) does offer a few tools for identifying and
auto-correcting errors in a dataset, but, as the name suggests, the
focus is on internal validity rather than general data cleaning. In
practice, this means that quite elegant tools for, e.g., linear
restraints among the variables in a dataset can be applied, but
looking for potentially miscoded missing values is not really
feasible. The main difference between these two challenges is the
direction in which the data is inspected: While linear constraints
work row-wise with no ambiguity, determining whether or not something
is a miscoded missing value often requires knowledge about the full
variable (e.g. range or data type), and thus it should be performed
column-wise. \pkg{validate} does not allow for user-defined extensions
of the latter type, thereby limiting its data cleaning potential.
Automatic data correction functions are also provided by
\pkg{validate} which we consider to be quite a dangerous cocktail: all
power is given to the the computer with no human supervision, and
investigators are less likely to make an active, case-specific choice
regarding the handling of the potential errors.  Finally, no tools
have been made to easily document exactly which checks and preliminary
results were used in the data cleaning process.

@@

\begin{table}[htbp]
\footnotesize
\bgroup
\def\arraystretch{1.8}%
\begin{tabular}{L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}}
& \code{summaryFunction} & \code{visualFunction} & \code{checkFunction} \\
\hline
\vspace{0pt} Input (required) & \code{v} --- a variable vector.
                                \newline \code{...} --- additional
                                arguments passed to the function.  &  \code{v} --- a variable vector. \newline \code{vnam} --- the variable name (as character string). \newline \code{doEval} --- a logical (\code{TRUE}/\code{FALSE}) controlling the output type of the function. & \code{v} - a variable vector \newline \code{nMax} - an integer (or \code{Inf}), controlling how many problematic values are printed, if relevant \newline \code{...} --- additional arguments passed to the function. \\
Input (optional) &  \code{maxDecimals} --- number of decimals printed in outputted numerical values.  & --- &  \code{maxDecimals}  --- number of decimals printed in outputted numerical values.  \\
Purpose & Describe some aspect of the variable, e.g., a central value, its dispersion or level of missingness. & Produce a distribution plot. & Check a variable for a specific issue and, if relevant, identify the values in the variable that cause the issue. \\
Output (required) & A list with entries: \newline \code{\$feature} --- a label for the summary value (as character string); \newline \code{\$result} --- the result of the summary (as character string). & A character string with \proglang{R} code for producing a plot. This code should be standalone, i.e., should include the data if necessary. & A list with entries: \newline \code{\$problem} --- a logical identifying whether an issue was found; \newline \code{\$message} --- a character string (possibly empty) decribing the issue that was found, properly escaped and ready for use with \code{rmarkdown}. \\
Output (recommended) & A \code{summaryResult} object, i.e., an attributed list with entries \code{\$feature}, \code{\$result} and \code{\$value}, the latter being the values from \code{\$result} in their original format). & \bgroup
\def\arraystretch{1.8}%
\hspace*{-.2cm}\begin{tabular}[t]{p{0.45\linewidth} |
                 p{0.45\linewidth}} \vspace*{-1cm}\raggedright If
                 \code{doEval} is \code{TRUE}: \newline A plot that is
                 opened by the graphic device in \proglang{R}. &
                                                          \vspace*{-1cm}\raggedright If \code{doEval} is \code{FALSE}: \newline   A text string with \proglang{R} code, as described above. \end{tabular} \egroup  & A \code{checkResult} object, i.e., an attributed list with entries \code{\$problem}, \code{\$message} and \code{\$problemValues}, the latter being either \code{NULL} or the problem causing values, as they were found in \code{v}, whichever is relevant. \\
Tools available for producing the function & \code{summaryResult()} & --- & \code{messageGenerator()} \newline \code{checkResult()}
\end{tabular}
\egroup
\caption{Reference information for creating new functions to be used
  as part of the summarize, visualize, and check steps. The three
  columns correspond to each of the three function types. %\hl{I think the formatting here is still sort of chaotic. Maybe include horizontal lines everywhere? Is it allowed in JSS? It's impossible to read what's in which row right now.}
}
\label{table.functionTypes}
\end{table}

@@

\subsubsection{Writing a summaryFunction}
As mentioned above, \pkg{dataMaid} provides a dedicated class for
\code{summaryFunction}s. However, this does not imply that they are
particularly advanced or complicated to create; in fact, they are
nothing but regular functions with a particular
input/output-structure. Specifically, they all follow the template
below:
\begin{Verbatim}
mySummaryFunction <- function(v, ...) {
  res <- [result of whatever summary we are doing]
  summaryResult(list(feature = "[Feature name]", result = res))
}
\end{Verbatim}
The last function called here, \code{summaryResult()}, changes the
class of the output, thereby making a \code{print()} method available
for it.  Note that \code{v} is an input vector and that \code{res}
should be either a character string or something that will be printed
as one. In other words an integer would be allowed, but a matrix will
not. Though a lot of different things can go into the
\code{summaryFunction} template, we recommend only using it for
summarizing the features of a variable, and leaving tests and checks
for the \code{checkFunction}s (presented below).

Adhering to the template above is sufficient for using the freshly
made \code{mySummaryFunction()} in \code{clean()}, but we recommend
adding the new function to the overview of all summary
functions by converting it to a proper \code{summaryFunction}
object. This is done by calling the \code{summaryFunction()} creator with
the user-defined function as the first argument, and additional arguments
\code{description} (an explanatory text which will be added to the attributes of
the function), and \code{classes} (a vector of variable classes the
user-defined function is intended to be applied to, also stored as an
attribute). In other words, a call following the template below should be made:
\begin{Verbatim}
mySummaryFunction <- summaryFunction(mySummaryFunction,
  description = "[Text describing what the summaryFunction does]",
  classes = c([vector of data types that the function is intended
  for]))
\end{Verbatim}
which adds the new function to the output of an
\code{allSummaryFunctions()} call.
% One comment should be devoted to
% the two attributes of a \code{summaryFunction}.
If the \code{description} argument is left unspecified, the name of
the function (which in this case is \code{"mySummaryFunction"}) will
be filled in and used as the \code{description} attribute. What
happens if the \code{classes} argument is not specified depends on the
type of \code{summaryFunction}. If \code{mySummaryFunction} is
constructed as an \code{S3} generic function with associated methods,
the call to \code{summaryFunction()} will automatically produce a
vector of the names of the classes for which the function can be
called. If \code{mySummaryFunction()} is not an \code{S3} generic and
\code{classes} is left unspecified, the attribute will simply be left
empty. Note that the helper function \code{allClasses()} might be
useful for filling out the \code{classes} argument, as it simply lists
all available classes in \pkg{dataMaid}:

@@

\subsubsection{Writing a visualFunction}
\code{visualFunction}s are the functions that produce the figures in a
\pkg{dataMaid} output document. Writing a \code{visualFunction} is slightly
more complicated than writing a \code{summaryFunction}. This follows from
the fact that \code{visualFunction}s need to be able to output standalone
code for plots in order for \code{clean()} to build standalone
\pkg{rmarkdown} files. We recommend using the following structure
(again shown as pseudo code):
\begin{Verbatim}
myVisualFunction <- function(v, vnam, doEval) {
  thisCall <- call("[the name of the function used to produce the plot]",
    v, [additional arguments for the plotting function])
  if (doEval) {
    return(eval(thisCall))
  } else return(deparse(thisCall))
}
\end{Verbatim}

In this function, \code{v} is the variable to be visualized,
\code{vnam} is its name (which should generally be passed to
\code{title} or \code{main} arguments in plotting functions) and
\code{doEval} controls whether the output is a plot (if \code{TRUE})
or a character string of standalone code for producing a plot (if
\code{FALSE}). Implementing the \code{doEval = TRUE} setting is not
strictly necessary for a \code{visualFunction}'s use in \code{clean()}, but
it makes it easier to assess what visualization options are available,
and obviously, it is crucial for interactive usage of
\code{myVisualFunction()}. In either case, it should be noted that all
the parameters listed above, \code{v}, \code{vnam} and \code{doEval},
are mandatory, so they must be left as is, even if they are not in
use (c.f., Table~\ref{table.functionTypes}).

@@

\subsubsection{Writing a checkFunction}
The last, but perhaps most important, \pkg{dataMaid} function type is
the \code{checkFunction}. These are the functions that flag issues in
the data as part of the check step and/or control the overall flow of the data
cleaning process in the pre-check stage. A \code{checkFunction} follows
one of two overall structures, depending on the type of check. Either,
it tries to identify problematic values in the variable (as
e.g., \code{identifyMissing()} does), or it performs a check concerning the
variable as a whole (e.g., the functions used for pre-checks and the
function \code{identifyNums()}). We present templates for both types of
\code{checkFunction}s below separately, but it should be emphasized that
formally, they belong to the same class.

@@

If problematic values are to be identified, the template from above
should be expanded to follow a slightly more complicated structure
(again shown as pseudo-code):

@@

One comment should be devoted to the helper function,
\code{messageGenerator()}.  This function's sole purpose is aiding
consistent styling of all \code{checkFunction} messages. The function
simply pastes together the \code{problemMessage} and the
\code{problemValues}, with the latter being quoted and sorted
alphabetically. If the \code{nMax} argument to \code{messageGenerator()} is
not \code{Inf}, only the first \code{nMax} problem values will be pasted
onto the message, accompanied by a comment about how many problem
values were left out (if any).  Note that printing quotes in
\proglang{R} markdown requires an extensive amount of character escaping, so
opting for \code{messageGenerator()} really is the easiest solution.

@@

\begin{Verbatim}
myProbValCheckFunction <- function(v, nMax, maxDecimals, ...) {
  [... more lines of code here ...]
  problemStatus <- list(problem = problem,
    problemValues = problemValues)

  if (!is.null(problemValues)) {
    problemStatus$problemValues <- round(problemValues, maxDecimals)
  }
  [... more lines of code here ...]
}
\end{Verbatim}
% $

Now, problematic values will be rounded in the outputted message, while they
will still appear in their original format under the entry \code{\$problemValues} %$
in the returned object.

@@

As an example, we create four new functions and show both how they can be
used interactively and how they can be integrated with the \code{clean()}
function. These four new functions are:
\begin{description}
\item[\code{isID()}] A new \code{checkFunction} intended for use in the pre-check-stage. This function checks whether a variable consists exclusively of long ($> 10$ characters/digits) entries that are all of equal length, as this might be personal identification codes that we do not wish to print out in the data summary.
\item[\code{mosaicVisual()}] A new \code{visualFunction} that produces mosaic plots. This function will be used in the \textit{visualize} step of \code{clean()}.
\item[\code{countZeros()}] A new \code{summaryFunction} that counts the number of occurrences of the value \code{0} in a variable. This function will be used in the \textit{summarize} step of \code{clean()}.
\item[\code{identifyColons()}] A new \code{checkFunction} that flags variables in which values have colons that appear before and after alphanumerical characters. This is practical for identifying autogenerated interaction effects. This function will be used in the \textit{check} step of \code{clean()}.
\end{description}
These functions are defined in turn below, and afterwards, an example of how they can be called from \code{clean()} is provided.

\subsubsection[isID --- a new checkFunction without problem values]{\code{isID} --- a new \code{checkFunction} without problem values}
First, let's define the \code{isID()} function. As this function is
not supposed to list problematic values, it falls within the category
of \code{checkFunction}s represented by
\code{myFullVarCheckFunction()} above. We do not particularly wish to
use this function interactively, so we will stick to the minimal
requirements of a \code{checkFunction} used in \code{check()} (see
Table~\ref{table.functionTypes}). The function can then be defined by

<<>>=
isID <- function(v, nMax = NULL, ...) {
  out <- list(problem = FALSE, message = "")
  if (class(v) %in% setdiff(allClasses(), c("logical", "Date"))) {
    v <- as.character(v)
    lengths <- c(nchar(v))
    if (all(lengths > 10) & length(unique(lengths)) == 1) {
      out$problem <- TRUE
      out$message <- "Warning: This variable seems to contain ID codes."
    }
  }
  out
}
@

This is essentially all we need to do in order to include this
function as a pre-check-function in \code{clean()}, so we will leave it as
is and move on to the next function, namely \code{mosaicVisual()}.

\subsubsection[mosaicVisual --- a new visualFunction]{\code{mosaicVisual} --- a new \code{visualFunction}}
\proglang{R} provides a function, \code{mosaicplot()}, which produces
mosaic plots. We intend to use this existing high-level plotting
function as part of the new visualization function.  We will define
the new function such that it gets the full \pkg{dataMaid}
functionality. This can be done using the following code which sets up
the call using the existing function \code{mosaicplot()}.

<<>>=
mosaicVisual <- function(v, vnam, doEval) {
  thisCall <- call("mosaicplot", table(v), main = vnam, xlab = "")
  if (doEval) {
    return(eval(thisCall))
  } else return(deparse(thisCall))
}
@

@@

Even though \code{mosaicVisual()}, as written above, follows the style of a
\code{visualFunction}, it is not yet truly one and therefore, it will not
appear in an \code{allVisualFunctions()} call. In order to get this
functionality we need to change its object class. This can be done by
writing

@@

\subsubsection[countZeros --- a new summaryFunction]{\code{countZeros} --- a new \code{summaryFunction}}
This \code{summaryFunction} is defined in the following lines of code:

@@

As this function computes an integer (the number of zeros),
there is no difference between the elements \code{\$result} and
\code{\$value}. If, on the other hand, the result had been a character
string, extra formatting might be required in the \code{\$result} entry
(such as escaping of quotation marks), and in this scenario, the two
entries would have differed. As the result is returned as a
\code{summaryResult} object, a printing method is automatically called
when \code{countZeros()} is used interactively:

@@

\subsubsection[identifyColons --- a new checkFunction with problem values]{\code{identifyColons} --- a new \code{checkFunction} with problem values}
The last function mentioned above is \code{identifyColons()}. We
define it using the helper function \code{messageGenerator()} to
obtain a properly escaped message, and we use \code{checkResult()} to
make its output print neatly. In the code we use regular expressions
through the \code{gregexpr()} function to identify colons between two
words.

@@

Note, that for \code{checkFunction}s, the description will appear in
the document produced by \code{clean()} (in the \textit{Data cleaning
  summary} section of the output), so for this type of function, the
change of class is not only done for the sake of the
\code{allCheckFunctions()} output.

\subsubsection[Calling the new summarize/visualize/check functions from clean()]{Calling the new summarize/visualize/check functions from \code{clean()}}
Now, we are ready to use these new functions in a \code{clean()}
call. The extended \pkg{dataMaid} output document should have the
following modifications, relative to the standard \pkg{dataMaid} output:
\begin{itemize}
\item We want to add the new pre-check function, \code{isID()}, to the already existing pre-checks.
\item We wish to change the plot type for all variables to the new mosaic plot.
\item We want the new summary function, \code{countZeros()}, to be added to the summaries performed on all variable types but \code{Date} and \code{logical}.
\item We want the new check function, \code{identifyColons()}, to be added to the checks performed on \code{character}, \code{factor} and \code{labelled} variables.
\end{itemize}
These options are specified as follows:

@@

Finally, we present a few examples of how to make \pkg{dataMaid}
solve specific issues related to data cleaning. First, we discuss the
challenges related to cleaning large datasets, particularly in terms
of memory use and computation speed. Next, we show how \pkg{dataMaid}
can be used for problem-flagging. Lastly, we discuss how the
\pkg{dataMaid} output document can be included in other \proglang{R} markdown
documents as a way to produce clear and concise documentation of a
dataset. %\hl{I feel like there should be more topics here, but I'm all
%  out of ideas...}

\subsection{Cleaning large datasets}
If the dataset becomes very large, the standard use of \code{clean()}
outlined above might not be ideal. If there is a large number of
variables, creation of the \proglang{R} markdown document might be quite
slow, while a large number of observations will generally affect the
rendering time of the document. This issue is especially relevant to
Windows users, where memory usage is not very efficient. The problem is that older,
32-bit versions of Windows generally cannot access all physical RAM,
while newer 64-bit versions of Windows cannot access virtual memory. Therefore,
memory allocation issues becomes even more pressing under Windows.

Below we give a few practical examples of ways to deal with large
data, while wishing to still produce (potentially very long) data
cleaning overview documents. Note that the interactive tools of
\pkg{dataMaid} can be used as usual or sequentially in small subsets
of the large dataset, if no overview documents are needed.

@@

Of course, even less heavy plots might be achieved by writing new
\code{visualFunction}s, using the guidelines from Section
\ref{sec:functionTemplates}. For instance, a future extension of
\pkg{dataMaid} might be the inclusion of ASCII plots, as
e.g., represented in the \proglang{R} package \pkg{txtplot} \citep{txtplot}.

@@

Now only the checking results are printed, and only for variables
where problems were identified. An even more minimal output can be
generated by also leaving out the checking results --- then
\code{clean()} essentially just produces a document with a list of the variable names
that should be investigated further:
